New insights into the Weddell Sea ecosystem applying a quantitative network approach

### 2022-12-30 Manuscript/Manuscript.Rmd
Version to pdf

---
title: New insights into the Weddell Sea ecosystem applying a quantitative network approach
journal: "`r rticles::copernicus_journal_abbreviations(journal_name = 'communication')`"
author:
  - given_name: Tomás I.
    surname: Marina
    affiliation: "1, *"
    email: tomasimarina@gmail.com
  - given_name: Leonardo A.
    surname: Saravia
    affiliation: "1, *"
    email: arysar@gmail.com
  - given_name: Susanne
    surname: Kortsch
    affiliation: 2

correspongdingauthors: Tomás I. Marina (tomasimarina@gmail.com) and Leonardo A. Saravia (arysar@gmail.com)

affiliation:
  - code: 1
    address: Centro Austral de Investigaciones Científicas (CADIC-CONICET), Ushuaia, Argentina
  - code: 2
    address: Spatial Foodweb Ecology Group, Department of Agricultural Sciences, University of Helsinki, Finland
  - code: "*"
    address: These authors contributed equally to this work.
abstract: |
  Network approaches can shed light on the structure and stability of complex marine communities. In recent years, such approaches have been successfully applied to study polar ecosystems, improving our knowledge on how they might respond to ongoing environmental changes. The Weddell Sea is one of the most studied marine ecosystems outside the Antarctic Peninsula in the Southern Ocean. Yet, few studies consider the known complexity of the Weddell Sea food web, which in its current form comprises 490 species and 16041 predator-prey interactions. Here we analysed the Weddell Sea food web, focusing on the species and trophic interactions that underpin ecosystem structure and stability. We estimated the strength for each interaction in the food web, characterised species position in the food web using unweighted and weighted food web properties, and analysed species’ roles with respect to the stability of the food web. We found that the distribution of the interaction strength (IS) at the food web level is asymmetric, with many weak interactions and few strong ones. We detected a positive relationship between species mean IS and two unweighted properties (i.e., trophic level and the total number of interactions). We also found that only a few species possess key positions in terms of food web stability. These species are characterised by high mean IS, mid to high trophic level, relatively high number of interactions, and mid to low trophic similarity. In this study, we integrated unweighted and weighted food web information, enabling a more complete assessment of the ecosystem structure and function of the Weddell Sea food web. Our results provide new insights, which are important for the development of effective policies and management strategies, particularly given the ongoing initiative to implement a Marine Protected Area (MPA) in the Weddell Sea.

bibliography: ../WeddellSea.bib
running:
  title: New insights into the Weddell Sea food web
  author: Marina et al.
competinginterests: |
  The authors declare no competing interests.

availability:

codedata: |
    R code and data set used in this work is available at https://github.com/EcoComplex/WeddellSea

authorcontribution: |
  TIM and LAS: Conceptualization (lead); Data curation (lead); Formal analysis (lead); Methodology (lead); Coding (lead); Writing – original draft (lead); Writing – review and editing (lead). SK: Conceptualization (lead); Formal analysis (supporting); Methodology (supporting); Coding (supporting); Writing – original draft (supporting); Writing – review and editing (supporting).

acknowledgements: |
  We thank all persons that dedicated time to build and curate the dataset GATEWAy that enable us to carry out this work. We also thank the organizers of the different virtual workshops of the Weddell Sea and Droning Maud Land (WSDML) Regional Working Group for giving us the possibility to participate and contribute in the Special Issue.

output:
  rticles::copernicus_article: 
    highlight: NULL
    keep_tex: true
  bookdown::pdf_book:
    base_format: rticles::copernicus_article
editor_options: 
  markdown: 
    wrap: 72
---

@@

Food web analysis constitutes an important framework for understanding ecological community structure and for conserving biodiversity through ecosystem management [@Thompson2012]. Although topological food web analysis, which considers only the presence and absence of predator-prey interactions, provides important insights into the structure and functioning of ecological communities [e.g. @Pascual2006; @Kortsch2015; @Marina2018; @Cordone2020; @Rodriguez2022], more information on the nature of the trophic interactions is needed to effectively charaterise ecosystem dynamics and stability [e.g. @Kortsch2021; @Pecuchet2022]. This is a fundamental step for providing assemessments on ecosystem vulnerability to environmental pressures and for prioritising management actions. In this regard, quantifying the strength of trohic interactions and species' roles within the network are of paramount importance [@Carrara2015; @Allesina2015; @Nilsson2016; @Cirtwill2018a].

Estimating interaction strength (IS) in food webs allows differentiating the importance of species interactions. On the contrary, unweighetd food web representations give equal importantance to all interactions well-knowing that some species interactions are stronger than others and hence play a diffrent role for ecosystem functioning and stability. Both empirical and theoretical studies show that interactions strength distributions in food webs are assymetric [@Paine1992; @McCann1998; @Emmerson2004; @Wootton2005; @Kortsch2021], containing a few strong and many weak interactions. This asymmetric patterning of weak and strong links is crucial to food web stability [@Paine1992; @McCann1998; @Neutel2002]. In a recent paper on an aquatic food web it was further highlighted that temporal changes in ecosytem functioning could only be predicted using weighted food web structure [@Kortsch2021]. Hence, in order to assess the stability and functioning of a food web, it is important to first determine the quantitative structure of the trophic network.

Several methodologies have been applied to estimate IS in food webs, where the quantity and quality of the data mostly determines which approach is the most convenient [@Berlow2004]. Approaches include experimental methods combined with dynamic modelling [@Emmerson2004; @Carrara2015], measurements of species abundances through time [@Fahimipour2014; @Chang2021], and estimation of metabolic rates and biomass of all species in the community [@Neutel2014]. However, these types of methods require large experimental set-ups and parameterisations restricting the analyses to smaller networks (e.g., approximatley 10 species or less). Other methods based on allometric scaling relationships and biomass information [@Kortsch2021; @Gauzens2019] can be applied to larger networks with less data requirements, but this comes at the expense of precision in the predictions. For even larger food webs composed of nearly 1000 species and more than 10000 interactions, only methods with even less data requirements are feasible. One of these methods, proposed by @Pawar2012, combines data on consumer and resource body masses and consumer search space (interaction dimensionality) to obtain IS estimates for each pairwise predator-prey interaction. An advantage of this method is that it can be applied without information on species biomass.

Using a a network approach, different types (e.g., terrestial, lake, marine) of food webs from various geographic locations are studied worldwide, including marine polar food webs [@Carscallen2012; @Santana2013; @Kortsch2019; @Pecuchet2022]. Some of the studies from the Arctic show how food web properties (e.g., connectance) are constrained by environmental factors such as sea ice cover and seawater temperature [@Kortsch2019; @Pecuchet2022]. In the Southern Ocean, important insights have been gained into mechanisms of energy flow, the relative importance of individual species and their traits, and the influence of environmental variables (e.g., sea-ice) on the structure of local food webs [@Cordone2020; @Rossi2019]. For instance, in Potter Cove (West Antarctic Peninsula) the substratum type (i.e. hard/soft or rocks/sediments) plays a significant role in the structure and stability of the food web. In Terra Nova Bay (Ross Sea), the architecture of biodiversity was reshaped by the pulsed input of sympagic food sources following sea-ice break up, with food web simplification, decreased intraguild predation, potential disturbance propagation and increased vulnerability to biodiversity loss [@Rossi2019].

The Weddell Sea is expected to be one of the last regions of the Southern Ocean to experience the consequences of climate change due to its extensive ice cover and ocean currents [@Teschke2021] resulting in less sea surface warming compared to other areas of the Southern Ocean. This Sea plays an important role in driving global thermohaline circulation and ventilating the global abyssal ocean because it generates a considerable part of the Antarctic Bottom Water [@Fahrbach2009]. Because of these environmental characteristics, the Weddell Sea may serve as a refuge for Antarctic species which depend on sea ice (e.g. krill, emperor penguin, Weddell seal) or have low heat tolerance (e.g., most notothenioid fishes) due to their adaptations to freezing tempeartures [@Griffiths2017]. While essential large-scale hydrodynamic relationships are relatively well-known for this region [@deSteur2019], information on the current distribution, abundance and sensitivity to climate change is only partially known for a few species (e.g., emperor penguin) [@Houstin2022].

>>>>

---
title: New insights into the Weddell Sea ecosystem applying a quantitative network approach
journal: "`r rticles::copernicus_journal_abbreviations(journal_name = 'communication')`"
author:
  - given_name: Tomás I.
    surname: Marina
    affiliation: "1, *"
    email: tomasimarina@gmail.com
  - given_name: Leonardo A.
    surname: Saravia
    affiliation: "1, *"
    email: arysar@gmail.com
  - given_name: Susanne
    surname: Kortsch
    affiliation: 2
correspongdingauthors: Tomás I. Marina (tomasimarina@gmail.com) and Leonardo A. Saravia (arysar@gmail.com)
affiliation:
  - code: 1
    address: Centro Austral de Investigaciones Científicas (CADIC-CONICET), Ushuaia, Argentina
  - code: 2
    address: Spatial Foodweb Ecology Group, Department of Agricultural Sciences, University of Helsinki, Finland
  - code: "*"
    address: These authors contributed equally to this work.
abstract: |
  Network approaches can shed light on the structure and stability of complex marine communities. In recent years, such approaches have been successfully applied to study polar ecosystems, improving our knowledge on how they might respond to ongoing environmental changes. The Weddell Sea is one of the most studied marine ecosystems outside the Antarctic Peninsula in the Southern Ocean. Yet, few studies consider the known complexity of the Weddell Sea food web, which in its current form comprises 490 species and 16041 predator-prey interactions. Here we analysed the Weddell Sea food web, focusing on the species and trophic interactions that underpin ecosystem structure and stability. We estimated the strength for each interaction in the food web, characterised species position in the food web using unweighted and weighted food web properties, and analysed species’ roles with respect to the stability of the food web. We found that the distribution of the interaction strength (IS) at the food web level is asymmetric, with many weak interactions and few strong ones. We detected a positive relationship between species mean IS and two unweighted properties (i.e., trophic level and the total number of interactions). We also found that only a few species possess key positions in terms of food web stability. These species are characterised by high mean IS, mid to high trophic level, relatively high number of interactions, and mid to low trophic similarity. In this study, we integrated unweighted and weighted food web information, enabling a more complete assessment of the ecosystem structure and function of the Weddell Sea food web. Our results provide new insights, which are important for the development of effective policies and management strategies, particularly given the ongoing initiative to implement a Marine Protected Area (MPA) in the Weddell Sea.

bibliography: ../WeddellSea.bib
running:
  title: New insights into the Weddell Sea food web
  author: Marina et al.
competinginterests: |
  The authors declare no competing interests.
availability:
codedata: |
    R code and data set used in this work is available at https://github.com/EcoComplex/WeddellSea
authorcontribution: |
  TIM and LAS: Conceptualization (lead); Data curation (lead); Formal analysis (lead); Methodology (lead); Coding (lead); Writing – original draft (lead); Writing – review and editing (lead). SK: Conceptualization (lead); Formal analysis (supporting); Methodology (supporting); Coding (supporting); Writing – original draft (supporting); Writing – review and editing (supporting).
acknowledgements: |
  We thank all persons that dedicated time to build and curate the dataset GATEWAy that enable us to carry out this work. We also thank the organizers of the different virtual workshops of the Weddell Sea and Droning Maud Land (WSDML) Regional Working Group for giving us the possibility to participate and contribute in the Special Issue.
output:
  rticles::copernicus_article: 
    highlight: NULL
    keep_tex: true
  bookdown::pdf_book:
    base_format: rticles::copernicus_article
editor_options: 
  markdown: 
    wrap: 72
---

@@

Food web analysis constitutes an important framework for understanding ecological community structure and for conserving biodiversity through ecosystem management [@Thompson2012]. Although topological food web analysis, which considers only the presence and absence of predator-prey interactions, provides important insights into the structure and functioning of ecological communities [e.g. @Pascual2006; @Kortsch2015; @Marina2018; @Cordone2020; @Rodriguez2022], more information on the nature of the trophic interactions is needed to effectively characterise ecosystem dynamics and stability [e.g. @Kortsch2021; @Pecuchet2022]. This is a fundamental step for providing assessments on ecosystem vulnerability to environmental pressures and for prioritising management actions. In this regard, quantifying the strength of trophic interactions and species' roles within the network are of paramount importance [@Carrara2015; @Allesina2015; @Nilsson2016; @Cirtwill2018a].

Estimating interaction strength (IS) in food webs allows differentiating the importance of species interactions. On the contrary, unweighted food web representations give equal importance to all interactions well-knowing that some species interactions are stronger than others and hence play a different role for ecosystem functioning and stability. Both empirical and theoretical studies show that interactions strength distributions in food webs are assymmetric [@Paine1992; @McCann1998; @Emmerson2004; @Wootton2005; @Kortsch2021], containing a few strong and many weak interactions. This asymmetric patterning of weak and strong links is crucial to food web stability [@Paine1992; @McCann1998; @Neutel2002]. In a recent paper on an aquatic food web it was further highlighted that temporal changes in ecosystem functioning could only be predicted using weighted food web structure [@Kortsch2021]. Hence, in order to assess the stability and functioning of a food web, it is important to first determine the quantitative structure of the trophic network.

Several methodologies have been applied to estimate IS in food webs, where the quantity and quality of the data mostly determines which approach is the most convenient [@Berlow2004]. Approaches include experimental methods combined with dynamic modelling [@Emmerson2004; @Carrara2015], measurements of species abundances through time [@Fahimipour2014; @Chang2021], and estimation of metabolic rates and biomass of all species in the community [@Neutel2014]. However, these types of methods require large experimental set-ups and parameterisations restricting the analyses to smaller networks (e.g., approximately 10 species or less). Other methods based on allometric scaling relationships and biomass information [@Kortsch2021; @Gauzens2019] can be applied to larger networks with less data requirements, but this comes at the expense of precision in the predictions. For even larger food webs composed of nearly 1000 species and more than 10000 interactions, only methods with even less data requirements are feasible. One of these methods, proposed by @Pawar2012, combines data on consumer and resource body masses and consumer search space (interaction dimensionality) to obtain IS estimates for each pairwise predator-prey interaction. An advantage of this method is that it can be applied without information on species biomass.

Using a a network approach, different types (e.g., terrestrial, lake, marine) of food webs from various geographic locations are studied worldwide, including marine polar food webs [@Carscallen2012; @Santana2013; @Kortsch2019; @Pecuchet2022]. Some of the studies from the Arctic show how food web properties (e.g., connectance) are constrained by environmental factors such as sea ice cover and seawater temperature [@Kortsch2019; @Pecuchet2022]. In the Southern Ocean, important insights have been gained into mechanisms of energy flow, the relative importance of individual species and their traits, and the influence of environmental variables (e.g., sea-ice) on the structure of local food webs [@Cordone2020; @Rossi2019]. For instance, in Potter Cove (West Antarctic Peninsula) the substratum type (i.e. hard/soft or rocks/sediments) plays a significant role in the structure and stability of the food web. In Terra Nova Bay (Ross Sea), the architecture of biodiversity was reshaped by the pulsed input of sympagic food sources following sea-ice break up, with food web simplification, decreased intraguild predation, potential disturbance propagation and increased vulnerability to biodiversity loss [@Rossi2019].

The Weddell Sea is expected to be one of the last regions of the Southern Ocean to experience the consequences of climate change due to its extensive ice cover and ocean currents [@Teschke2021] resulting in less sea surface warming compared to other areas of the Southern Ocean. This Sea plays an important role in driving global thermohaline circulation and ventilating the global abyssal ocean because it generates a considerable part of the Antarctic Bottom Water [@Fahrbach2009]. Because of these environmental characteristics, the Weddell Sea may serve as a refuge for Antarctic species which depend on sea ice (e.g. krill, emperor penguin, Weddell seal) or have low heat tolerance (e.g., most notothenoid fishes) due to their adaptations to freezing temperatures [@Griffiths2017]. While essential large-scale hydrodynamic relationships are relatively well-known for this region [@deSteur2019], information on the current distribution, abundance and sensitivity to climate change is only partially known for a few species (e.g., emperor penguin) [@Houstin2022].
